Variance reduction in coarse bifurcation analysis of stochastic models

### 2016-07-04 Documentation/talk_21_06.tex
last bifurcation results

\title[Systemic Risk] % (optional, only for long titles)
{Variance Reduction in Equation-free Newton-Krylov-Methods}
%\subtitle{Evidence from India}
%\author[Author, Anders] % (optional, for multiple authors)
%{F.~Author\inst{1} \and S.~Anders\inst{2}}
%\institute[Universities Here and There] % (optional)
%{
%  \inst{1}%
%  Institute of Computer Science\\
%  University Here
%  \and
%  \inst{2}%
%  Institute of Theoretical Philosophy\\
%  University There
%}
\date{July 12, 2016} % (optional)
%{Conference on Presentation Techniques, 2004}
%\subject{Computer Science}

@@

\tableofcontents

\section{Model Problem}

\begin{frame}
\frametitle{Fokker-Planck equation}

\begin{equation*}
\label{fokkerplanck}
\pa{\rho(x,t)}{t} + \pa{(\mu(x) \rho(x,t))}{x} = \frac{\sigma^2}{2}  \ppa{\rho(x,t)}{x} 
\end{equation*}
%\begin{center}
%%\includegraphics[width=6.5cm]{../Problems/WeightedParticles/checkSystem/plots/rho_pde}
%\end{center}

Simulate an ensemble of $N$ particles evolving according to the corresponding SDE
\begin{equation*}
    \dd X = \mu f(X) \dd t + \sigma  \dd{W_t}.
\end{equation*}

\begin{frame}

\frametitle{Coarse time stepper}

\begin{equation*}
\U(t + n \Delta t) = \cts(\U) = (\mathcal{R} \circ \mathcal{E}(n \Delta t, \omega) \circ  \mathcal{L}(\mathbf{\omega})  ) (\U(t))
\end{equation*}

\begin{itemize} [<+->]
\item 
\begin{block} 
{Lifting $\mathcal{L}: \U \rightarrow \mathbf{x}  $}
\end{block}
\item
 \begin{block} {Evolution  $\mathcal{E}$: \\ Simulation of the SDE for $N$ particles over $n$ timesteps. }
\begin{equation*}
   \mathbf{X}^{n + 1} = \mathbf{X}^{n} + \mu (\mathbf{X}^n) \Delta t +  \sqrt{\sigma \Delta t}\cdot \boldsymbol{\xi}^n 
\end{equation*}
with \begin{equation*}\xi^n(\omega)  \sim \mathcal{N} (0,1)
\end{equation*}
\end{block}

\item 
\begin{block} {Restriction $\mathcal{R}: \mathbf{x} \rightarrow \U $}
\begin{equation*}
\frac{1}{N} \sum_{i=1}^{N}  w_i \cdot \chi_{\Delta_j}(X_i) = \rho_j  
\end{equation*}
\end{block}

\begin{frame}
\frametitle{Evaluating the bias on Jacobian-vector products}

\begin{equation*} 
 \mathbf{Bias}(\jv, \jvpde ) = \mathbb{\hat{E}}[\jv]  - \jvpde 
\end{equation*}

with 
\begin{equation*}
\jvpde \approx \frac{ \U^{\varepsilon}(t + n \Delta t)  - \U (t +  n\Delta t)}{\varepsilon}
\end{equation*}
 \pause

and $\U$ calculated by explicitly solving

\begin{equation*}
\label{fokkerplanck}
\pa{\rho(x,t)}{t} + \pa{(\mu(x) \rho(x,t))}{x} = D  \ppa{\rho(x,t)}{x} 
\end{equation*}

 using

\begin{equation*} 
\rho_i^{n+1} = \rho_i^n + \Delta t \left( \frac{D}{{\Delta x}^2} \left( \rho_{i+1}^{n} - 2 \rho_i^n + \rho_{i-1}^n \right)  - \frac{a(x)}{\Delta x} (\rho_i^n - \rho_{i-1}^n) \right)
\end{equation*}
for the value of $\rho$ at position $x=i \Delta x$ and time $t=(n+1) \Delta t$

\begin{frame}
\frametitle{Evaluating the bias on Jacobian-vector products}

\begin{equation*} 
 \mathbf{Bias}(\jv, \jvpde ) = \mathbb{\hat{E}}[\jv]  - \jvpde 
\end{equation*}

 \begin{figure}
 \includegraphics[width=9.5cm]{../Problems/WeightedParticles/checkSystem/plots/Bias_Jv_N_M.pdf}

\begin{frame}
\frametitle{Evaluating the bias on Jacobian-vector products}

\begin{equation*} 
 \mathbf{Bias}(\jv, \jvpde ) = \mathbb{\hat{E}}[\jv]  - \jvpde 
\end{equation*}

 \begin{figure}
 \includegraphics[width=9.5cm]{../Problems/WeightedParticles/checkSystem/plots/Bias_Jv_M_N.pdf}

@@

@@

\begin{equation*}
\label{Jv_approx}
J(\cts) \cdot \mathbf{v} \approx \frac{\cts (\U + \varepsilon \mathbf{v}, \boldsymbol{\omega_1} )  - \cts (\U, \boldsymbol{\omega_2})}{\varepsilon} 
%&\approx & \frac{\cts (\U, \boldsymbol{\omega_1} )  + \varepsilon D(\cts) (  \U, \boldsymbol{\omega_1})  \cdot \mathbf{v}  - \cts (\U, \boldsymbol{\omega_2}) }{\varepsilon} \nonumber
\end{equation*}

%\pause
%\item Problem: $ \texttt{Var}(\jv)  \sim \mathcal{O}(1/(\varepsilon^2  N)) $\\$ \Rightarrow $     Numerical noise for $ \varepsilon \ll 1$    
%%\end{frame}
%\end{itemize}

\pause
\begin{block}{Problem}
 $ \texttt{Var}(\jv)  \sim \mathcal{O}(1/(\varepsilon^2  N)) \Rightarrow $     Numerical noise for $ \varepsilon \ll 1$ 
 \end{block}  
%\end{frame}
\end{itemize}
\end{frame}
%
%\begin{frame} \frametitle{Variance reduction of Jacobian-vector products}
%
%\begin{equation*}
%\label{Jv_approx}
%J(\cts) \cdot \mathbf{v} \approx \frac{\cts (\U + \varepsilon \mathbf{v}, \boldsymbol{\omega_1} )  - \cts (\U, \boldsymbol{\omega_2})}{\varepsilon} 
%%&\approx & \frac{\cts (\U, \boldsymbol{\omega_1} )  + \varepsilon D(\cts) (  \U, \boldsymbol{\omega_1})  \cdot \mathbf{v}  - \cts (\U, \boldsymbol{\omega_2}) }{\varepsilon} \nonumber
%\end{equation*}
%Problem: $ \texttt{Var}(\jv)  \sim \mathcal{O}(1/(\varepsilon^2  N)) \Rightarrow $ 
%    Numerical noise for $ \varepsilon \ll 1$    
%\end{frame}
%

@@

\section{Research Plan}

\section{Other}

\begin{frame} {Courses}

\begin{block}{Contribution to Education}
\begin{itemize}
\item Exercises Analysis 1
\item Exercises Analysis 2
\end{itemize}
\end{block}

\begin{block}{Doctoral Training Programme}
\begin{itemize}
\item Seminar scientific integrity 
\item Teacher assistant training 
\item Science Communication and Outreach 
\end{itemize}
\end{block}

\begin{block}{Other courses}
\begin{itemize}
\item Functional Analysis
\end{itemize}
\end{block}
\end{frame}

>>>>

 % % %

\title[Systemic Risk] % (optional, only for long titles)
{Applying Newton-Krylov Solver on a Model for Systemic Risk}
%\subtitle{Evidence from India}
%\author[Author, Anders] % (optional, for multiple authors)
%{F.~Author\inst{1} \and S.~Anders\inst{2}}
%\institute[Universities Here and There] % (optional)
%{
%  \inst{1}%
%  Institute of Computer Science\\
%  University Here
%  \and
%  \inst{2}%
%  Institute of Theoretical Philosophy\\
%  University There
%}
\date{\today} % (optional)
%{Conference on Presentation Techniques, 2004}
%\subject{Computer Science}

@@

%\tableofcontents

@@

%\begin{frame}
%
%\frametitle{Coarse time stepper}
%
%\begin{equation*}
%\U(t + n \Delta t) = \cts(\U) = (\mathcal{R} \circ \mathcal{E}(n \Delta t, \omega) \circ  \mathcal{L}(\mathbf{\omega})  ) (\U(t))
%\end{equation*}
%
%
%\begin{itemize} [<+->]
%\item 
%\begin{block} 
%{Lifting $\mathcal{L}: \U \rightarrow \mathbf{x}  $}
%\end{block}
%\item
% \begin{block} {Evolution  $\mathcal{E}$: \\ Simulation of the SDE for $N$ particles over $n$ timesteps. }
%\begin{equation*}
%   \mathbf{x}^{n + 1} = \mathbf{x}^{n} + a(\mathbf{x}^n) \Delta t +  \sqrt{2D \Delta t}\cdot \boldsymbol{\xi}^n 
%\end{equation*}
%with \begin{equation*}\xi^n(\omega)  \sim \mathcal{N} (0,1)
%\end{equation*}
%\end{block}
%
%\item 
%\begin{block} {Restriction $\mathcal{R}: \mathbf{x} \rightarrow \U $}
%\begin{equation*}
%\frac{1}{N} \sum_{i=1}^{N}  w_i \cdot \chi_{\Delta_j}(x_i) = \rho_j  
%\end{equation*}
%\end{block}fjlk
%
%\end{itemize}
%
%\end{frame}
%
%
%\begin{frame}
%\frametitle{Evaluating Jacobian-vector products}
%\begin{equation*}
% \U_* - \cts(\U_*) =0
%\end{equation*}
%
%\begin{equation*}
%\mathbf{Jv} = D(\cts) \cdot \mathbf{v} \approx \frac{\cts (\U + \varepsilon \mathbf{v}, \boldsymbol{\omega_1} )  - \cts (\U, \boldsymbol{\omega_2})}{\varepsilon} 
%\end{equation*}
%
%\pause
%\begin{block}{Problem: numerical noise for $ \varepsilon \ll 1$ }
%Repeating $\cts$ with two sets of random numbers $\boldsymbol{\omega_1},  \boldsymbol{\omega_2}$  will give different results
%$\rightarrow  \mathcal{O}(1/(\varepsilon^2 N))$ variance. 
%\end{block}
%
%\end{frame}
%

@@

\end{itemize}
\begin{equation*}
\label{Jv_approx}
J(\cts) \cdot \mathbf{v} \approx \frac{\cts (\U + \varepsilon \mathbf{v}, \boldsymbol{\omega_1} )  - \cts (\U, \boldsymbol{\omega_2})}{\varepsilon} 
%&\approx & \frac{\cts (\U, \boldsymbol{\omega_1} )  + \varepsilon D(\cts) (  \U, \boldsymbol{\omega_1})  \cdot \mathbf{v}  - \cts (\U, \boldsymbol{\omega_2}) }{\varepsilon} \nonumber
\end{equation*}
%Problem: $ \texttt{Var}(\jv)  \sim \mathcal{O}(1/(\varepsilon^2  N)) \Rightarrow $ 
%    Numerical noise for $ \varepsilon \ll 1$    
%\end{frame}
\end{frame}
%
%\begin{frame} \frametitle{Variance reduction of Jacobian-vector products}
%
%\begin{equation*}
%\label{Jv_approx}
%J(\cts) \cdot \mathbf{v} \approx \frac{\cts (\U + \varepsilon \mathbf{v}, \boldsymbol{\omega_1} )  - \cts (\U, \boldsymbol{\omega_2})}{\varepsilon} 
%%&\approx & \frac{\cts (\U, \boldsymbol{\omega_1} )  + \varepsilon D(\cts) (  \U, \boldsymbol{\omega_1})  \cdot \mathbf{v}  - \cts (\U, \boldsymbol{\omega_2}) }{\varepsilon} \nonumber
%\end{equation*}
%Problem: $ \texttt{Var}(\jv)  \sim \mathcal{O}(1/(\varepsilon^2  N)) \Rightarrow $ 
%    Numerical noise for $ \varepsilon \ll 1$    
%\end{frame}
%

@@
